### 2016-05-04 p03-ch02-2014-jmseq/jmseq-elsa.tex
Change bib style to 'american' from the original template
Fix ref/label for Chapter 05

 \section{Introduction} \label{sec:intro}
Monitoring refers to the activity of tracking observable information from an
execution of a system with the goal of checking it against the constraints imposed
by the system specification. The observable information of the monitored system
typically includes behavior, input and output, but may also contain quantitative
information. A monitoring framework consists of a monitor that extracts the
observable behavior from a program execution, and a controller that checks the
behavior against a set of specifications. Should an execution
violate the constraints imposed by the specification, corrective
actions can be taken.

@@

\begin{figure}[t]
\begin{center}
\includegraphics[scale=0.5]{images/arch-diagram-summary-exec}
\caption{JMSeq Execution Architecture}
\label{fig:arch-overview}
\end{center}
\end{figure}

In this paper, we introduce JMSeq, a tool for monitoring  sequences of method calls in Java programs.
JMSeq is based on Java annotations rather than (source code or bytecode) instrumentation. 
%, at source or bytecode level, we use another approach, called JMSeq.
%JMSeq monitors  sequences of method calls using.
Therefore, it does \emph{not} require  access to the source code of the target application.
JMSeq detects the annotated methods from the target application at runtime. 
Furthermore,  Java annotations themselves have no direct effect on program execution or logic of the application;
they carry metadata that can optionally be used at runtime to perform operations orthogonal to the execution of the main application.
   %Annotations are based on a declarative language to specifiy method call sequences using regular expressions for method signatures.
%The program is verified against the specifications during runtime and fails as soon as a violation is detected. 
%The verification is performed in a separate virtual machine to protect the original application from any modification (Figure \ref{fig:arch-overview}). 
%In the following, we provide a  discussion of what distinguishes JMSeq from other related work (c.f. Section \ref{sec:relworks}) in verification:
The main distinguishing overall characteristic of JMSeq is
the integration of the following main features.

\begin{description}
%   \item[Specification language.] JMSeq introduces a simple declarative specification language.
%   The programmer can use the language to express verification requirements using simple expressions from the language.
%   The language is based on regular expressions of method call signatures.
%   The language is simple and intuitive to help the programmer identify different types of sequences of method calls.
   \item[Java annotations.] 
%JMSeq uses Java annotations to specify and verify the sequence of method calls. 
JMSeq annotations specify  method call sequences  in terms of regular expressions of method call signatures.
 This pure declarative language is simple and intuitive in  specification  of  different types of sequences of method calls.
   \item[JVM decoupling.]
To ensure that the target program is in \emph{no} way affected or changed during monitoring, JMSeq runs in a separate JVM from the target application's JVM.
   JMSeq uses Java Platform Debugger API (JPDA) \cite{JPDA_Home} to initiate a separate JVM and verify the target application using events from the target JVM.
  Figure \ref{fig:arch-overview} displays the high-level execution architecture of JMSeq.
   \item[Fail-fast approach.] JMSeq terminates execution as soon as a violation of a specification occurs.  This approach prevents unsafe behavior,
   enables the localization of an error, and facilitates a reaction
   to a failure immediately, thus making the software more robust~\cite{Shore04a}.
   %In mission-critical software, this approach is favored over others because it facilitates a reaction to the situation as soon as it occurs.
   %It helps the software/service provider to be able to provide better software to its end users.
   %It additionally contributes to better disaster recovery of software in failure situation.
\end{description}

An important consequence of these features
is that neither  the source code (even if available) nor bytecode is modified by JMSeq. 
Furthermore,  JMSeq does not require source code to be present.  
As such, JMSeq is suited for
the specification and verification of both
 newly built software and legacy systems, including third-party libraries
(and even the Java standard library).
For example, given  the API of a legacy system, its expected behavior  can be
specified by annotating the methods of a ``driver'' program which triggers
the legacy code.
JMSeq then verifies the triggered behavior of the legacy code using these
annotations.
%
%   In the context of legacy systems, it can be assumed that the sequence of API and method calls are known.
%   It is also known that JMSeq annotations are not placed on the legacy system.
%   This means that a helper program is needed to use the annotations from JMSeq.
%   Therefore, one can write a helper program that uses the legacy system and its API.
%   The helper program is annotated with JMSeq annotations.
%   Then, the helper program is executed through JMSeq.
%   JMSeq verifies the legacy system through the helper program that carries the annotations and as such the specification of method call sequences from the legacy system.
%   JMSeq verification of legacy system is distinguished from approaches that use instrumentation (especially source level) because access to source code cannot be presumed.
 In the context of an ongoing software development process, JMSeq can also be used as mentioned above
by means of a ``driver'' program but also by directly
annotating the ``interfaces'' of the system API.
%   Since Java annotations have no effect on the execution of the application, they behave as metadata at runtime.
%   JMSeq can detect them and verify the system without the need for a helper program.
By annotating the  interfaces of a system any service object that implements the interface behavior inherits the annotations and can be verified by JMSeq.
In this  respect, JMSeq also introduces integration with frameworks such as JUnit (cf. Section \ref{sec:junit}).

The rest of the paper is organized as follows.
The first three sections describe the usage of JMSeq.
Section \ref{sec:spec} presents the language for specifying sequences of method calls, 
and Section \ref{sec:impl_annot} discusses the types of annotations available in JMSeq.
These two sections are connected via examples in Section \ref{sec:samples}.
Afterwards, implementation of the JMSeq framework is explained in detail in Section \ref{sec:implementation}.
The practical applicability of JMSeq is demonstrated by means of a case study from industry in Section \ref{sec:frh} and performance evaluation in Section \ref{sec:results}. 
Finally, related work is discussed in Section \ref{sec:relworks}.
Section \ref{sec:conclusion} concludes the paper. %discuss possible future work.

\section{Method Call Sequence Specification} \label{sec:spec}

We consider a component to be a collection of compiled Java classes. The
relevant dynamic behavior of a component can be expressed in terms of specific
sequences of messages among a small, fixed number of objects. In
Figure \ref{fig:seq-spec}, we see two UML message sequence diagrams each
describing how, and in which order, the four objects interact with each other. In this
section, we develop a specification language for describing these kinds of
interactions.

@@

In addition to specifying the order, a specification language for sequences of method calls
needs to distinguish between the following two cases (depicted in
Figure \ref{fig:seq-spec}).
\begin{description}
  \item[Case $1$] 
  shows a scenario in which (the call to) \texttt{m\_c} is nested in 
  \texttt{m\_b} since \texttt{m\_c} is called during the activation of 
  \texttt{m\_b} (i.e., after \texttt{m\_b} is called and before it returns). 
  Similarly, both \texttt{m\_b} and \texttt{m\_c} are nested in \texttt{m\_a}.

@@

  \item[Case $2$] represents a method call in which methods from
  different/same objects are called in a sequential rather than nested
  manner. In this example, both \texttt{m\_b} and \texttt{m\_c} are called by
  \texttt{m\_a}.
\end{description}
Typically, a program will need a combination of both cases to specify its dynamic behavior.
Specifying only the order of the method calls is not enough, as both cases above
have the same order of method calls. It is thus required to have a specification
technique that distinguishes between the \textsl{method calls} and
\textsl{method returns}. \begin{figure}[h]
\begin{center}
  \includegraphics[scale=0.4]{images/seq-spec}
  \caption{Examples of Method Call Sequence Specification}
  \label{fig:seq-spec}
\end{center}
\end{figure}

@@

In JMSeq, a specification is written as a post-condition associated with a method. It
specifies the set of possible sequences of method calls, or protocol, of an
object in the context of the relevant part of its environment. Formally,
sequences of method calls can be specified by means
of the grammar in  Figure \ref{fig:spec_grammar}.

\begin{figure}[h]
\begin{center}
\begin{framed}
\includegraphics[scale=1]{grammar}
\end{framed}
\caption{Method Sequence Specification Grammar}
\label{fig:spec_grammar}
\end{center}
\end{figure}

@@

Next we give a few examples of correct method sequence specifications. 
For instance, the specification of the sequence in case $1$ of
Figure \ref{fig:seq-spec} is given by:
%\begin{flushleft}
%\includegraphics[scale=1]
\input{grammar-sample-usage-1-1}
%\end{flushleft}
%\begin{alltt}
%\centering\small
%{\color{red}\{}call(* *.A.m_a(..)){\color{red}[}{\color{blue}\{}call(*
%*.B.m_b(..)){\color{blue}[}\{call(*
%*.C.m_c(..))\}{\color{blue}]}{\color{blue}\}}{\color{red}]}{\color{red}\}}
%\end{alltt}
Here it is important to notice that the call to method  \texttt{m\_b} is internal
to \texttt{m\_a}, and the call to \texttt{m\_c} is internal to \texttt{m\_b}.
The sequence in case $2$ of the same figure would be:
%\begin{flushleft}
%\includegraphics[scale=1]
\input{grammar-sample-usage-2-1}
%\end{flushleft}
%\begin{alltt}
%\centering\small
%{\color{red}\{}call(* *.A.m_a(..)){\color{red}[}\{call(*
%*.B.m_b(..))\}{\color{red}]}{\color{red}[}\{call(*
%*.C.m_c(..))\}{\color{red}]}{\color{red}\}}
%\end{alltt}
where both calls to methods  \texttt{m\_b} and \texttt{m\_c} are internal to
\texttt{m\_a}.

@@

Such a specification is used in circumstances where the execution of  \texttt{m\_a} is a prerequisite to the execution of either of \texttt{m\_b} or \texttt{m\_c}.
% 
% It is \emph{notable} that the meta-grammar provided in Figure
% \ref{fig:spec_grammar} is a context-free grammar; 
It is notable that the actual specifications are regular expressions;
for example, they may contain  unbounded repetitions and a choice of calls.

\section{Annotations for method call sequences} \label{sec:impl_annot}
Java enables developers to attach extra metadata to their programs using annotations.
%Java provides a means to annotate parts of a program.
These metadata do not directly affect the program semantics, but they do affect the
way programs are treated by tools and libraries, which can in turn affect the
behavior of the running program \cite{Java_5_Annotations}. Annotations can be
read from source files, class files, or reflectively at runtime. 
One can define custom annotation types and use them to provide the desired types of metadata for the declarations in his program. 
In Java, an annotation is a special kind of modifier, and can be used anywhere other
modifiers (such as {\small\texttt{public}}, {\small\texttt{static}}, or
{\small\texttt{final}}) can be used.

We use annotations as an enabling feature to test components without the need to have their source code.
We only use the metadata loaded from the annotations during runtime.
These metadata include the specification of the sequences of method calls
based on the grammar discussed in Section \ref{sec:spec}. The metadata
is placed on classes or methods in a Java program. 
Then, they are loaded into JMSeq when it monitors the actual observable
behavior of the program at runtime.

@@

To verify the sequence of executions, methods in classes need to be
annotated. Thus, JMSeq needs to detect all  classes that contain
some method carrying an annotation for verification. 
To improve readability and performance, JMSeq introduces
\smalltt{SequencedObject} annotation. This annotation is a marker
for such classes and notifies the annotation metadata loader (cf.
Section \ref{sec:annotation_repo}) that the objects instantiated from the annotated
class contain methods that specify a sequence of method executions. 
The code is given in Listing \ref{lst:seq_obj}.

@@

In Listing~\ref{lst:seq_obj}, {\small\texttt{@Retention(RetentionPolicy.RUNTIME)}} declares that this
annotation is only applicable during runtime and may  not be used in other scenarios.
Moreover, {\small\texttt{@Target(ElementType.TYPE)}} declares that this
annotation can only be used on types including classes, interfaces and
enumerated types; in other words, this annotation may not be used to mark a method
in a Java class.

\subsection{Sequenced Method Annotations}
A sequenced method annotation is used to specify the sequences of method
calls beginning from a given method. The annotation requires a string
property declaring the sequence specification discussed in Section
\ref{sec:spec}. Listing \ref{lst:seq_method} presents the declaration of
this annotation type.

@@

{\small\texttt{@Target(ElementType.METHOD)}} declares that this annotation is
only applicable to \textit{methods}. The string value from
{\small\texttt{value()}} holds the sequence specification (cf.
Section \ref{sec:spec}). 
Additionally, a sequenced method annotation allows verification of
exceptions during runtime. The tester can configure whether exceptions
are allowed or not using the combination of \smalltt{allowExceptions()}
and \smalltt{expect()}:
\begin{description}
  \item[\smalltt{allowExceptions()=true}] implies that exceptions are allowed during execution; 
  this is the default case.
If there is no exception configured using \smalltt{expect()},
\emph{any} exception is acceptable; otherwise,  an exception is acceptable 
only if it is specified in the array
of classes in \smalltt{expect()} or if it is a subclass of one of them.
If an exception occurs and it is ``expected'', JMSeq allows the program to continue execution;
the program should nevertheless be able to properly deal with the exception.

@@

JMSeq additionally provides a general mechanism to specify how to handle situations in
which the sequenced execution fails to complete. This can happen
because of a method call which was not included in the specification, or
an unexpected exception.
To this end, JMSeq introduces \smalltt{Verification\-FailureHandler} interface as an
extension point where the tester or developer can provide 
custom behavior to gracefully deal with the verification failure.
This interface
introduces a single method: \smalltt{handle\-FailedExecution(Execution
e)}. 
This method
%\smalltt{handle\-FailedExecution(Execution e)} 
receives information about the execution through a parameter of
type \smalltt{Execution} (cf. Section \ref{sec:implementation}).
\smalltt{Execution} is the central data model that carries information
about the method execution sequences; the trail of the
executions may be useful to the developer to examine the failure causes.
%
If no such handler is provided, verification failure results in termination 
of program execution with a failure status.

Section \ref{sec:samples} provides more discussion along with different
examples on using JMSeq.

\section{JMSeq by example} \label{sec:samples}
In this section, we go through different usage scenarios of JMSeq 
by means of illustrative examples.

\subsection{Sequenced Execution Specification}
In  Listing \ref{lst:annotated_spec}, we again go through the example
introduced in Section \ref{sec:spec}.
% give two examples of
% annotations of the class {\small\texttt{Main}} in Figure
% \ref{fig:seq-spec}. 
In this example, we annotate the method \texttt{main()} with the sequences of method calls describing the two scenarios
given in Figure \ref{fig:seq-spec}. 
The JMSeq specifications used here are discussed in Section \ref{sec:spec}.
The listing shows how to use them in a Java program.

@@

\subsection{Exception Verification}
We discuss how JMSeq can be used to verify  occurrence of exceptions during
execution of an application using the example given in Figure
\ref{fig:seq-spec} Case 1.

Let's assume that \smalltt{m\_c} may encounter
\smalltt{NullPointerException} and \smalltt{m\_b} may face
\smalltt{IOException}. But, we do not hold any assumption on the
location in the code where the exceptions may be handled or even
totally ignored. We use the following annotation in Listing
\ref{lst:case-1-exceptions} to express this situation.

@@

JMSeq additionally supports integration with testing frameworks such as JUnit. 
An example of such integration is discussed in Section \ref{sec:junit}.

\section{The JMSeq framework} \label{sec:implementation}

@@

As discussed above, to use JMSeq, it is assumed that Java components are supplied with sequenced object and sequenced method annotations. 
Multiple annotations are possible within the same class, but not for the same method. 
This implies that a component specification is in fact scattered across its constituent classes, each of which with a local and partial view of the component behavior.
Only the annotated methods will be monitored by JMSeq.
The overall verification process in JMSeq is depicted in Figure \ref{fig:alg-verify}.

\begin{figure*}[t]
\begin{center}
  \includegraphics[scale=0.47]{images/alg-diagram}
\end{center}
\caption{Overview of JMSeq's process to verify a program}
\label{fig:alg-verify}
\end{figure*}

@@

Before verification starts, JMSeq inspects the classes in the execution class path searching for methods that are annotated for sequenced verification; during this process, it collects all of the available method call sequence metadata. 
This step (denoted as Load Annotated Metadata in Figure \ref{fig:alg-verify}) does not need the source code, but rather uses the annotations metadata available in the compiled code.

@@

The process of matching method execution events with state machine transitions is done in JMSeq by following these steps:
\begin{enumerate}
  \item %A ``call expression'' is constructed based on the information associated to the received event. 
  First, every event received from the target JVM, together with the information associated to the event, is transformed to an instance of \smalltt{Execution} to capture various details about the execution of a method. 
  Then, at verification time, the instance of \smalltt{Execution} is transformed to an instance of \smalltt{CallExpression} to hold method signature details used by the state machine to make a transition. (Section \ref{sec:event_handling})
  \item Using the metadata available from the annotations of the methods, then, the   next ``possible call expressions'' of the current state, corresponding to the possible next transitions of the state machine,  are built. (Section \ref{sec:annotation_repo}, \ref{sec:exec_verify})
  \item A match making is done between the possible call expressions (state machine transitions) and the   current call expression as the candidate.  
  If a match is found,  the transition is made and the method is  accepted; otherwise, it fails. 
  When a failure occurs, JMSeq will execute the custom verification failure handler that is implemented either as part of
  the component code by the programmer, or externally by the system tester. (Section \ref{sec:exec_verify})
\end{enumerate}

@@

The overall architecture of JMSeq is given in Figure \ref{fig:arch-diag}. It basically
consists of three main modules: one for handling the events raised by the JVM executing the target program, another module for storing the annotation information, and a third
one to perform the runtime verification.

\begin{figure*}[t]
\begin{center}
  \includegraphics[scale=0.35]{images/arch-diagram}
  \caption{Software Architecture for Method Sequence Specification}
  \label{fig:arch-diag}
\end{center}
\end{figure*}

% The current design of JMSeq is completely \textsl{general} and \textsl{modular};  as it
% allows for replacing the grammar in Figure \ref{fig:spec_grammar} with other
% specification modules, based, for example, on temporal logics or extended
% regular expressions.

\subsubsection{Event Handling Module} \label{sec:event_handling}
Java Debugger Interface (JDI) is the interface
of Java Platform Debugger Architecture (JPDA) that gives access to different details during the execution of a program.
Following the JDI event model, JMSeq takes control over some of the
execution events that JVM publishes during a program execution. Therefore, a
component was designed to model and hold the execution trace events required
for event handling and execution verification.

>>>>

 \section{Introduction} \label{ch05:sec:intro}
Monitoring refers to the activity of tracking observable information from an
execution of a system with the goal of checking it against the constraints imposed
by the system specification. The observable information of the monitored system
typically includes behavior, input and output, but may also contain quantitative
information. A monitoring framework consists of a monitor that extracts the
observable behavior from a program execution, and a controller that checks the
behavior against a set of specifications. Should an execution
violate the constraints imposed by the specification, corrective
actions can be taken.

@@

\begin{figure}[t]
\begin{center}
\includegraphics[scale=0.5]{images/arch-diagram-summary-exec}
\caption{JMSeq Execution Architecture}
\label{ch05:fig:arch-overview}
\end{center}
\end{figure}

In this paper, we introduce JMSeq, a tool for monitoring  sequences of method calls in Java programs.
JMSeq is based on Java annotations rather than (source code or bytecode) instrumentation. 
%, at source or bytecode level, we use another approach, called JMSeq.
%JMSeq monitors  sequences of method calls using.
Therefore, it does \emph{not} require  access to the source code of the target application.
JMSeq detects the annotated methods from the target application at runtime. 
Furthermore,  Java annotations themselves have no direct effect on program execution or logic of the application;
they carry metadata that can optionally be used at runtime to perform operations orthogonal to the execution of the main application.
   %Annotations are based on a declarative language to specifiy method call sequences using regular expressions for method signatures.
%The program is verified against the specifications during runtime and fails as soon as a violation is detected. 
%The verification is performed in a separate virtual machine to protect the original application from any modification (Figure \ref{ch05:fig:arch-overview}). 
%In the following, we provide a  discussion of what distinguishes JMSeq from other related work (c.f. Section \ref{ch05:sec:relworks}) in verification:
The main distinguishing overall characteristic of JMSeq is
the integration of the following main features.

\begin{description}
%   \item[Specification language.] JMSeq introduces a simple declarative specification language.
%   The programmer can use the language to express verification requirements using simple expressions from the language.
%   The language is based on regular expressions of method call signatures.
%   The language is simple and intuitive to help the programmer identify different types of sequences of method calls.
   \item[Java annotations.] 
%JMSeq uses Java annotations to specify and verify the sequence of method calls. 
JMSeq annotations specify  method call sequences  in terms of regular expressions of method call signatures.
 This pure declarative language is simple and intuitive in  specification  of  different types of sequences of method calls.
   \item[JVM decoupling.]
To ensure that the target program is in \emph{no} way affected or changed during monitoring, JMSeq runs in a separate JVM from the target application's JVM.
   JMSeq uses Java Platform Debugger API (JPDA) \cite{JPDA_Home} to initiate a separate JVM and verify the target application using events from the target JVM.
  Figure \ref{ch05:fig:arch-overview} displays the high-level execution architecture of JMSeq.
   \item[Fail-fast approach.] JMSeq terminates execution as soon as a violation of a specification occurs.  This approach prevents unsafe behavior,
   enables the localization of an error, and facilitates a reaction
   to a failure immediately, thus making the software more robust~\cite{Shore04a}.
   %In mission-critical software, this approach is favored over others because it facilitates a reaction to the situation as soon as it occurs.
   %It helps the software/service provider to be able to provide better software to its end users.
   %It additionally contributes to better disaster recovery of software in failure situation.
\end{description}

An important consequence of these features
is that neither  the source code (even if available) nor bytecode is modified by JMSeq. 
Furthermore,  JMSeq does not require source code to be present.  
As such, JMSeq is suited for
the specification and verification of both
 newly built software and legacy systems, including third-party libraries
(and even the Java standard library).
For example, given  the API of a legacy system, its expected behavior  can be
specified by annotating the methods of a ``driver'' program which triggers
the legacy code.
JMSeq then verifies the triggered behavior of the legacy code using these
annotations.
%
%   In the context of legacy systems, it can be assumed that the sequence of API and method calls are known.
%   It is also known that JMSeq annotations are not placed on the legacy system.
%   This means that a helper program is needed to use the annotations from JMSeq.
%   Therefore, one can write a helper program that uses the legacy system and its API.
%   The helper program is annotated with JMSeq annotations.
%   Then, the helper program is executed through JMSeq.
%   JMSeq verifies the legacy system through the helper program that carries the annotations and as such the specification of method call sequences from the legacy system.
%   JMSeq verification of legacy system is distinguished from approaches that use instrumentation (especially source level) because access to source code cannot be presumed.
 In the context of an ongoing software development process, JMSeq can also be used as mentioned above
by means of a ``driver'' program but also by directly
annotating the ``interfaces'' of the system API.
%   Since Java annotations have no effect on the execution of the application, they behave as metadata at runtime.
%   JMSeq can detect them and verify the system without the need for a helper program.
By annotating the  interfaces of a system any service object that implements the interface behavior inherits the annotations and can be verified by JMSeq.
In this  respect, JMSeq also introduces integration with frameworks such as JUnit (cf. Section \ref{ch05:sec:junit}).

The rest of the paper is organized as follows.
The first three sections describe the usage of JMSeq.
Section \ref{ch05:sec:spec} presents the language for specifying sequences of method calls, 
and Section \ref{ch05:sec:impl_annot} discusses the types of annotations available in JMSeq.
These two sections are connected via examples in Section \ref{ch05:sec:samples}.
Afterwards, implementation of the JMSeq framework is explained in detail in Section \ref{ch05:sec:implementation}.
The practical applicability of JMSeq is demonstrated by means of a case study from industry in Section \ref{ch05:sec:frh} and performance evaluation in Section \ref{ch05:sec:results}. 
Finally, related work is discussed in Section \ref{ch05:sec:relworks}.
Section \ref{ch05:sec:conclusion} concludes the paper. %discuss possible future work.

\section{Method Call Sequence Specification} \label{ch05:sec:spec}

We consider a component to be a collection of compiled Java classes. The
relevant dynamic behavior of a component can be expressed in terms of specific
sequences of messages among a small, fixed number of objects. In
Figure \ref{ch05:fig:seq-spec}, we see two UML message sequence diagrams each
describing how, and in which order, the four objects interact with each other. In this
section, we develop a specification language for describing these kinds of
interactions.

@@

In addition to specifying the order, a specification language for sequences of method calls
needs to distinguish between the following two cases (depicted in
Figure \ref{ch05:fig:seq-spec}).
\begin{description}
  \item[Case $1$] 
  shows a scenario in which (the call to) \texttt{m\_c} is nested in 
  \texttt{m\_b} since \texttt{m\_c} is called during the activation of 
  \texttt{m\_b} (i.e., after \texttt{m\_b} is called and before it returns). 
  Similarly, both \texttt{m\_b} and \texttt{m\_c} are nested in \texttt{m\_a}.

@@

  \item[Case $2$] represents a method call in which methods from
  different/same objects are called in a sequential rather than nested
  manner. In this example, both \texttt{m\_b} and \texttt{m\_c} are called by
  \texttt{m\_a}.
\end{description}
Typically, a program will need a combination of both cases to specify its dynamic behavior.
Specifying only the order of the method calls is not enough, as both cases above
have the same order of method calls. It is thus required to have a specification
technique that distinguishes between the \textsl{method calls} and
\textsl{method returns}. \begin{figure}[h]
\begin{center}
  \includegraphics[scale=0.4]{images/seq-spec}
  \caption{Examples of Method Call Sequence Specification}
  \label{ch05:fig:seq-spec}
\end{center}
\end{figure}

@@

In JMSeq, a specification is written as a post-condition associated with a method. It
specifies the set of possible sequences of method calls, or protocol, of an
object in the context of the relevant part of its environment. Formally,
sequences of method calls can be specified by means
of the grammar in  Figure \ref{ch05:fig:spec_grammar}.

\begin{figure}[h]
\begin{center}
\begin{framed}
\includegraphics[scale=1]{grammar}
\end{framed}
\caption{Method Sequence Specification Grammar}
\label{ch05:fig:spec_grammar}
\end{center}
\end{figure}

@@

Next we give a few examples of correct method sequence specifications. 
For instance, the specification of the sequence in case $1$ of
Figure \ref{ch05:fig:seq-spec} is given by:
%\begin{flushleft}
%\includegraphics[scale=1]
\input{grammar-sample-usage-1-1}
%\end{flushleft}
%\begin{alltt}
%\centering\small
%{\color{red}\{}call(* *.A.m_a(..)){\color{red}[}{\color{blue}\{}call(*
%*.B.m_b(..)){\color{blue}[}\{call(*
%*.C.m_c(..))\}{\color{blue}]}{\color{blue}\}}{\color{red}]}{\color{red}\}}
%\end{alltt}
Here it is important to notice that the call to method  \texttt{m\_b} is internal
to \texttt{m\_a}, and the call to \texttt{m\_c} is internal to \texttt{m\_b}.
The sequence in case $2$ of the same figure would be:
%\begin{flushleft}
%\includegraphics[scale=1]
\input{grammar-sample-usage-2-1}
%\end{flushleft}
%\begin{alltt}
%\centering\small
%{\color{red}\{}call(* *.A.m_a(..)){\color{red}[}\{call(*
%*.B.m_b(..))\}{\color{red}]}{\color{red}[}\{call(*
%*.C.m_c(..))\}{\color{red}]}{\color{red}\}}
%\end{alltt}
where both calls to methods  \texttt{m\_b} and \texttt{m\_c} are internal to
\texttt{m\_a}.

@@

Such a specification is used in circumstances where the execution of  \texttt{m\_a} is a prerequisite to the execution of either of \texttt{m\_b} or \texttt{m\_c}.
% 
% It is \emph{notable} that the meta-grammar provided in Figure
% \ref{ch05:fig:spec_grammar} is a context-free grammar; 
It is notable that the actual specifications are regular expressions;
for example, they may contain  unbounded repetitions and a choice of calls.

\section{Annotations for method call sequences} \label{ch05:sec:impl_annot}
Java enables developers to attach extra metadata to their programs using annotations.
%Java provides a means to annotate parts of a program.
These metadata do not directly affect the program semantics, but they do affect the
way programs are treated by tools and libraries, which can in turn affect the
behavior of the running program \cite{Java_5_Annotations}. Annotations can be
read from source files, class files, or reflectively at runtime. 
One can define custom annotation types and use them to provide the desired types of metadata for the declarations in his program. 
In Java, an annotation is a special kind of modifier, and can be used anywhere other
modifiers (such as {\small\texttt{public}}, {\small\texttt{static}}, or
{\small\texttt{final}}) can be used.

We use annotations as an enabling feature to test components without the need to have their source code.
We only use the metadata loaded from the annotations during runtime.
These metadata include the specification of the sequences of method calls
based on the grammar discussed in Section \ref{ch05:sec:spec}. The metadata
is placed on classes or methods in a Java program. 
Then, they are loaded into JMSeq when it monitors the actual observable
behavior of the program at runtime.

@@

To verify the sequence of executions, methods in classes need to be
annotated. Thus, JMSeq needs to detect all  classes that contain
some method carrying an annotation for verification. 
To improve readability and performance, JMSeq introduces
\smalltt{SequencedObject} annotation. This annotation is a marker
for such classes and notifies the annotation metadata loader (cf.
Section \ref{ch05:sec:annotation_repo}) that the objects instantiated from the annotated
class contain methods that specify a sequence of method executions. 
The code is given in Listing \ref{ch05:lst:seq_obj}.

@@

In Listing~\ref{ch05:lst:seq_obj}, {\small\texttt{@Retention(RetentionPolicy.RUNTIME)}} declares that this
annotation is only applicable during runtime and may  not be used in other scenarios.
Moreover, {\small\texttt{@Target(ElementType.TYPE)}} declares that this
annotation can only be used on types including classes, interfaces and
enumerated types; in other words, this annotation may not be used to mark a method
in a Java class.

\subsection{Sequenced Method Annotations}
A sequenced method annotation is used to specify the sequences of method
calls beginning from a given method. The annotation requires a string
property declaring the sequence specification discussed in Section
\ref{ch05:sec:spec}. Listing \ref{ch05:lst:seq_method} presents the declaration of
this annotation type.

@@

{\small\texttt{@Target(ElementType.METHOD)}} declares that this annotation is
only applicable to \textit{methods}. The string value from
{\small\texttt{value()}} holds the sequence specification (cf.
Section \ref{ch05:sec:spec}). 
Additionally, a sequenced method annotation allows verification of
exceptions during runtime. The tester can configure whether exceptions
are allowed or not using the combination of \smalltt{allowExceptions()}
and \smalltt{expect()}:
\begin{description}
  \item[\smalltt{allowExceptions()=true}] implies that exceptions are allowed during execution; 
  this is the default case.
If there is no exception configured using \smalltt{expect()},
\emph{any} exception is acceptable; otherwise,  an exception is acceptable 
only if it is specified in the array
of classes in \smalltt{expect()} or if it is a subclass of one of them.
If an exception occurs and it is ``expected'', JMSeq allows the program to continue execution;
the program should nevertheless be able to properly deal with the exception.

@@

JMSeq additionally provides a general mechanism to specify how to handle situations in
which the sequenced execution fails to complete. This can happen
because of a method call which was not included in the specification, or
an unexpected exception.
To this end, JMSeq introduces \smalltt{Verification\-FailureHandler} interface as an
extension point where the tester or developer can provide 
custom behavior to gracefully deal with the verification failure.
This interface
introduces a single method: \smalltt{handle\-FailedExecution(Execution
e)}. 
This method
%\smalltt{handle\-FailedExecution(Execution e)} 
receives information about the execution through a parameter of
type \smalltt{Execution} (cf. Section \ref{ch05:sec:implementation}).
\smalltt{Execution} is the central data model that carries information
about the method execution sequences; the trail of the
executions may be useful to the developer to examine the failure causes.
%
If no such handler is provided, verification failure results in termination 
of program execution with a failure status.

Section \ref{ch05:sec:samples} provides more discussion along with different
examples on using JMSeq.

\section{JMSeq by example} \label{ch05:sec:samples}
In this section, we go through different usage scenarios of JMSeq 
by means of illustrative examples.

\subsection{Sequenced Execution Specification}
In  Listing \ref{ch05:lst:annotated_spec}, we again go through the example
introduced in Section \ref{ch05:sec:spec}.
% give two examples of
% annotations of the class {\small\texttt{Main}} in Figure
% \ref{ch05:fig:seq-spec}. 
In this example, we annotate the method \texttt{main()} with the sequences of method calls describing the two scenarios
given in Figure \ref{ch05:fig:seq-spec}. 
The JMSeq specifications used here are discussed in Section \ref{ch05:sec:spec}.
The listing shows how to use them in a Java program.

@@

\subsection{Exception Verification}
We discuss how JMSeq can be used to verify  occurrence of exceptions during
execution of an application using the example given in Figure
\ref{ch05:fig:seq-spec} Case 1.

Let's assume that \smalltt{m\_c} may encounter
\smalltt{NullPointerException} and \smalltt{m\_b} may face
\smalltt{IOException}. But, we do not hold any assumption on the
location in the code where the exceptions may be handled or even
totally ignored. We use the following annotation in Listing
\ref{ch05:lst:case-1-exceptions} to express this situation.

@@

JMSeq additionally supports integration with testing frameworks such as JUnit. 
An example of such integration is discussed in Section \ref{ch05:sec:junit}.

\section{The JMSeq framework} \label{ch05:sec:implementation}

@@

As discussed above, to use JMSeq, it is assumed that Java components are supplied with sequenced object and sequenced method annotations. 
Multiple annotations are possible within the same class, but not for the same method. 
This implies that a component specification is in fact scattered across its constituent classes, each of which with a local and partial view of the component behavior.
Only the annotated methods will be monitored by JMSeq.
The overall verification process in JMSeq is depicted in Figure \ref{ch05:fig:alg-verify}.

\begin{figure*}[t]
\begin{center}
  \includegraphics[scale=0.47]{images/alg-diagram}
\end{center}
\caption{Overview of JMSeq's process to verify a program}
\label{ch05:fig:alg-verify}
\end{figure*}

@@

Before verification starts, JMSeq inspects the classes in the execution class path searching for methods that are annotated for sequenced verification; during this process, it collects all of the available method call sequence metadata. 
This step (denoted as Load Annotated Metadata in Figure \ref{ch05:fig:alg-verify}) does not need the source code, but rather uses the annotations metadata available in the compiled code.

@@

The process of matching method execution events with state machine transitions is done in JMSeq by following these steps:
\begin{enumerate}
  \item %A ``call expression'' is constructed based on the information associated to the received event. 
  First, every event received from the target JVM, together with the information associated to the event, is transformed to an instance of \smalltt{Execution} to capture various details about the execution of a method. 
  Then, at verification time, the instance of \smalltt{Execution} is transformed to an instance of \smalltt{CallExpression} to hold method signature details used by the state machine to make a transition. (Section \ref{ch05:sec:event_handling})
  \item Using the metadata available from the annotations of the methods, then, the   next ``possible call expressions'' of the current state, corresponding to the possible next transitions of the state machine,  are built. (Section \ref{ch05:sec:annotation_repo}, \ref{ch05:sec:exec_verify})
  \item A match making is done between the possible call expressions (state machine transitions) and the   current call expression as the candidate.  
  If a match is found,  the transition is made and the method is  accepted; otherwise, it fails. 
  When a failure occurs, JMSeq will execute the custom verification failure handler that is implemented either as part of
  the component code by the programmer, or externally by the system tester. (Section \ref{ch05:sec:exec_verify})
\end{enumerate}

@@

The overall architecture of JMSeq is given in Figure \ref{ch05:fig:arch-diag}. It basically
consists of three main modules: one for handling the events raised by the JVM executing the target program, another module for storing the annotation information, and a third
one to perform the runtime verification.

\begin{figure*}[t]
\begin{center}
  \includegraphics[scale=0.35]{images/arch-diagram}
  \caption{Software Architecture for Method Sequence Specification}
  \label{ch05:fig:arch-diag}
\end{center}
\end{figure*}

% The current design of JMSeq is completely \textsl{general} and \textsl{modular};  as it
% allows for replacing the grammar in Figure \ref{ch05:fig:spec_grammar} with other
% specification modules, based, for example, on temporal logics or extended
% regular expressions.

\subsubsection{Event Handling Module} \label{ch05:sec:event_handling}
Java Debugger Interface (JDI) is the interface
of Java Platform Debugger Architecture (JPDA) that gives access to different details during the execution of a program.
Following the JDI event model, JMSeq takes control over some of the
execution events that JVM publishes during a program execution. Therefore, a
component was designed to model and hold the execution trace events required
for event handling and execution verification.
